Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
addd automaton todo
two related work todos
small changes to semantics section

To justify our assertion that \theoryabbv is a syntax for formal
grammars and parse transformers, we will now define a
\emph{denotational semantics} that makes this mathematically
precise. To interpret our types of codes we assume a universe of
\emph{small} sets.

@@

First, we need to establish a precise definition of a formal grammar
that accounts for the possibility of ambiguity. The most common
definition of a formal grammar is as generative grammars, defined by
a set of non-terminals, with a specified start symbol and set of
production rules. We instead use a more abstract formulation that is
closer in spirit to the standard definition of a formal language:

\begin{definition}
  A \emph{formal language} $L$ is a function from strings to propositions.
  A \emph{formal grammar} $A$ is a function from strings to small sets.
\end{definition}

We think of the grammar $A$ as taking a string to the set of all parse
trees for that string. However since $A$ could be any function
whatsoever there is no requirement that an element of $A(w)$ be a
``tree'' in the usual sense. This definition provides a simple,
syntax-independent definition of a grammar that can be used for any
formalism: generative grammars, categorial grammars, or our own
type-theoretic grammars. Note that the definition of a formal grammar
is a generalization of the usual notion of formal language since a
proposition can be equivalently defined as a subset of a one-element
set. Then the difference between a formal grammar and a formal
language is that formal grammars can be \emph{ambiguous} in that there
can be more than one parse of the same string. Even for unambiguous
grammars, we care not just about \emph{whether} a string has a parse
tree, but \emph{which} parse tree it has, i.e., what the structure of
the element of $A(w)$ is.

The linear types of \theoryabbv can all be interpreted as formal
grammars, but what about the linear terms? These can be interpreted as
\emph{parse transformers}, which we now define.
\begin{definition}
  Let $L_1,L_2$ be formal languages. Then we say $L_1\subseteq L_2$ if
  for every string $w$, $L_1(w)\Rightarrow L_2(w)$.

  Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
  $A_1$ to $A_2$ is a function assigning to each string $w$ a function
  $f_w : A_1(w) \to A_2(w)$.
\end{definition}
\pedro{Do we actually need a definition of formal language morphism?
In case we do, it should probably be its own definition.}
First, note that parse transformers generalize language inclusion: if
$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
In our denotational semantics, linear terms will be interpreted as
such parse transformers, and the notions of unambiguous grammar,
parseability, etc, introduced in \Cref{sec:concepts} can be verified
to correspond to their intended meanings under this interpretation.

Parse transformers can be composed: given two parse transformers $f$ and $g$,
their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
Furthemore, given a formal grammar $A$, its identity transformer is $id_w =
id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
defines a \emph{category}.
\begin{definition}
  Define $\Grammar$ to be the category whose objects are formal
  grammars and morphisms are parse transformers.
\end{definition}
This category is equivalent to the slice category $\Set/\Sigma^*$ and
as such is very well-behaved. It is complete, co-complete, Cartesian
closed and carries a monoidal biclosed structure. We will use these
structures to model the linear types, terms and equalities in
\theoryabbv.
%% \begin{theorem}
%%   The category $\Grammar$ is complete, co-complete and cartesian
%%   closed. Furthermore, it is monoidal biclosed with the monoidal
%%   product given by
%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
%%   and unit by
%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
%% \end{theorem}

@@

% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
% \steven{I am in favor of cutting what's above, but we haven't introduced the
%   syntax in the main body of the paper yet, even if the figures appeared
%   earlier. So what you write needs to take that into account}
% \pedro{In which case we should introduce it in the previous section. I suggest
%   adding this right after ``and the intuitionistic typing rules in fig 3''}
\subsection{Semantics}
\steven{Add definitions of kleene,lambek,and chomsky categories here?}

We can now define our denotational semantics.
\begin{definition}[Grammar Semantics]
  We define the following interpretations by mutual recursion on the
  judgments of \theoryabbv:
  \begin{enumerate}
  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
    similarly define a formal grammar $\sem\Delta\gamma$ for each
    linear context $\Gamma \Delta\isLinCtx$.
  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
  \end{enumerate}
  And we verify the following conditions:
  \begin{enumerate}
  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
  \end{enumerate}
\end{definition}

For reasons of brevity, we present only the cases related to the
linear types in \Cref{fig:semantics}. The full details are in the
appendix.\max{TODO: put the full details in the appendix}.  The
grammar for a literal $c$ has a single parse precisely when the input
string consists of the single character. The grammar for the unit
similarly has a single parse for the empty string.  A parse of the
tensor product $A \otimes B$ consists of a \emph{splitting} of the
empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
is a pair of an element of the set $X$ and a parse of $A(x)$, while
dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
that takes an $A$ parse of some other string $w'$ to a $B$ parse of
$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.

The universe $L$ of linear types is interpreted as the set of all
(small) grammars.

\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    \sem{L}\gamma &= \Grammar_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

The most complex part of the semantics is the interpretation of
strictly positive functors and indexed inductive linear types.  We
interpret a strictly positive functor as a \emph{dependent polynomial
functor} on the category of sets, also sometimes called an
\emph{indexed container} \cite{dep-poly,indexed-container}.
\begin{definition}
  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
  $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
  of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
  composite
  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXFNldC9JIl0sWzMsMCwiXFxTZXQvTyJdLFsxLDAsIlxcU2V0L1AiXSxbMiwwLCJcXFNldC9TIl0sWzAsMiwiZl4qIl0sWzIsMywiXFxQaV9nIl0sWzMsMSwiXFxTaWdtYV9oIl1d
  \[\begin{tikzcd}
	    {\Set/I} & {\Set/P} & {\Set/S} & {\Set/O}
	    \arrow["{f^*}", from=1-1, to=1-2]
	    \arrow["{\Pi_g}", from=1-2, to=1-3]
	    \arrow["{\Sigma_h}", from=1-3, to=1-4]
  \end{tikzcd}\]
  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the depedent
  product operation and $\Sigma_h$ is the dependent sum operation, which
  are, respectively, the right and left adjoint of their pullback functors
  $g^*$ and $h^*$.

  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
  \to \Set/O$ that is nautrally isomorphic to the extension of a
  dependent polynomial from $I$ to $O$.
\end{definition}
With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
as the action of the functor on objects and morphisms,
respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
functors in the obvious way that matches the definitional behavior of
$\el$ and $\map$. The non-trivial part of the construction is
verifying that such functors are constructing corresponding dependent
polynomials. The details are tedious but straightforward extension of
prior work on dependent polynomials and indexed containers and we have
verified the construction in Agda.

The reason we use dependent polynomials functors on sets is that it is
well-known that their initial algebras exist
\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
for dependent polynomials are straightforwardly constructed in Agda as
an inductive type of ``IW'' trees, which is available as a
construction in the cubical library of Agda \cite{iw,cubical-library}.
\begin{theorem}
  Every dependent polynomial functor from $I$ to $I$ has an initial
  algebra.
\end{theorem}
Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
$\sem{X}\gamma$-indexed family of polynomial functors from
$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
of these constructs a polynomial functor from $\sem{X}\gamma\times
\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
initial algebra of this functor, and the initial algebra structure is
used to interpret $\roll, \fold$ and the corresponding axioms.

@@

\section{Discussion and Future Work}
\label{sec:discussion}
\subsection{Typed Approaches to Grammars}
\subsubsection{Lambek Calculus and Categorial Grammar}

>>>>

To justify our assertion that \theoryabbv is a syntax for formal
grammars and parse transformers, we will now define a
\emph{denotational semantics} that makes this mathematically precise
by defining a notion of formal grammar and parse transformer and
showing that our type theory can be soundly interpreted in this
model. We then discuss how this denotational semantics provides the
basis for our prototype implementation in Agda.

@@

The most common definition of a formal grammar is as generative
grammars, defined by a set of non-terminals, a specified start symbol
and set of production rules. We instead use a more abstract
formulation that is closer in spirit to the standard definition of a
formal \emph{language}\cite{elliott}:
\begin{definition}
  A \emph{formal language} $L$ is a function from strings to propositions.
  A (small) \emph{formal grammar} $A$ is a function from strings to (small) sets.
\end{definition}
We think of the grammar $A$ as taking a string to the set of all parse
trees for that string. However since $A$ could be any function
whatsoever there is no requirement that an element of $A(w)$ be a
``tree'' in the usual sense. This definition provides a simple,
syntax-independent definition of a grammar that can be used for any
formalism: generative grammars, categorial grammars, or our own
type-theoretic grammars. Note that the definition of a formal grammar
is a generalization of the usual notion of formal language since a
proposition can be equivalently defined as a subset of a one-element
set. Then the difference between a formal grammar and a formal
language is that formal grammars can be \emph{ambiguous} in that there
can be more than one parse of the same string. Even for unambiguous
grammars, we care not just about \emph{whether} a string has a parse
tree, but \emph{which} parse tree it has, i.e., what the structure of
the element of $A(w)$ is.  To interpret our universes $U , L$ we
assume we have a universe of \emph{small} sets. In the remainder, all
formal grammars are assumed to be small.

We then interpret linear \emph{terms} as \emph{parse transformers}:
\begin{definition}
  Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
  $A_1$ to $A_2$ is a function assigning to each string $w$ a function
  $f_w : A_1(w) \to A_2(w)$.
\end{definition}
Just as formal grammars generalize formal languages, parse
transformers generalize formal language inclusion: if $A_1(w), A_2(w)$
are all subsets of a one-element set, then a parse transformer is
equivalent to showing that $A_1(w) \subseteq A_2(w)$. In our
denotational semantics, linear terms will be interpreted as such parse
transformers, and the notions of unambiguous grammar, parsers,
disjointness, etc, introduced in \Cref{sec:applications} can be
verified to correspond to their intended meanings under this
interpretation.

Parse transformers can be composed: given two parse transformers $f$ and $g$,
their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
Furthemore, given a formal grammar $A$, its identity transformer is $id_w =
id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
defines a \emph{category}.
\begin{definition}
  Define $\Grammar$ to be the category whose objects are formal
  grammars and morphisms are parse transformers.
\end{definition}
This category is equivalent to the slice category $\Set/\Sigma^*$ and
as such is very well-behaved. It is complete, co-complete, Cartesian
closed and carries a monoidal biclosed structure. We will use these
structures to model the linear types, terms and equalities in
\theoryabbv. These categorical properties are precisely what is
required to interpret all of the linear type and term constructors.
%% \begin{theorem}
%%   The category $\Grammar$ is complete, co-complete and cartesian
%%   closed. Furthermore, it is monoidal biclosed with the monoidal
%%   product given by
%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
%%   and unit by
%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
%% \end{theorem}

@@

% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
% \steven{I am in favor of cutting what's above, but we haven't introduced the
%   syntax in the main body of the paper yet, even if the figures appeared
%   earlier. So what you write needs to take that into account}
% \pedro{In which case we should introduce it in the previous section. I suggest
%   adding this right after ``and the intuitionistic typing rules in fig 3''}
\subsection{Semantics}

We now define our denotational semantics.
\begin{definition}[Grammar Semantics]
  We define the following interpretations by mutual recursion on the
  judgments of \theoryabbv:
  \begin{enumerate}
  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
    similarly define a formal grammar $\sem\Delta\gamma$ for each
    linear context $\Gamma \Delta\isLinCtx$.
  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
  \end{enumerate}
  And we verify the following conditions:
  \begin{enumerate}
  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
  \item If $\Gamma \vdash X \equiv X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
  \item If $\Gamma \vdash A \equiv A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
  \item If $\Gamma \vdash M \equiv M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
  \item If $\Gamma;\Delta \vdash e \equiv e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
  \end{enumerate}
\end{definition}

The interpretation of dependent types as sets is standard
\cite{dependent-types-as-sets}. We present the concrete descriptions
of the semantics of linear types, as well as our non-standard
non-linear types in \Cref{fig:semantics}.
The grammar for a literal $c$ has a single parse
precisely when the input string consists of the single character. The
grammar for the unit similarly has a single parse for the empty
string.  A parse of the tensor product $A \otimes B$ consists of a
\emph{splitting} of the empty string into a prefix $w_1$ and suffix
$w_2$ along with an $A$ parse of $w_1$ and $B$ parse of $w_2$. A parse
of $\bigoplus_{x:X} A$ is a pair of an element of the set $X$ and a
parse of $A(x)$, while dually a parse of $\bigwith_{x:X} A$ is a
\emph{function} taking any $x:X$ to a parse of $A(x)$. A $w$-parse of
$A \lto B$ is a function that takes an $A$ parse of some other string
$w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same except the
$B$ parse is for the reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.
%
The universe $L$ of linear types is interpreted as the set of all
(small) grammars.
\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    \sem{L}\gamma &= \Grammar_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

The most complex part of the semantics is the interpretation of
strictly positive functors and indexed inductive linear types.  We
interpret a strictly positive functor as a \emph{dependent polynomial
functor} on the category of sets, also sometimes called an
\emph{indexed container} \cite{dep-poly,indexed-container}.
\begin{definition}
  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
  $P$ and functions $f : P \to I$, $g : P \to S$ and $h : S \to O$. The \emph{extension}
  of a polynomial is a functor $\Set/I \to \Set/O$ defined as the
  composite
  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXFNldC9JIl0sWzMsMCwiXFxTZXQvTyJdLFsxLDAsIlxcU2V0L1AiXSxbMiwwLCJcXFNldC9TIl0sWzAsMiwiZl4qIl0sWzIsMywiXFxQaV9nIl0sWzMsMSwiXFxTaWdtYV9oIl1d
  \[\begin{tikzcd}
	    {\Set/I} & {\Set/P} & {\Set/S} & {\Set/O}
	    \arrow["{f^*}", from=1-1, to=1-2]
	    \arrow["{\Pi_g}", from=1-2, to=1-3]
	    \arrow["{\Sigma_h}", from=1-3, to=1-4]
  \end{tikzcd}\]
  Where $f^*$ is the pullback functor along $f$; $\Pi_g$ is the depedent
  product operation and $\Sigma_h$ is the dependent sum operation, which
  are, respectively, the right and left adjoint of their pullback functors
  $g^*$ and $h^*$. A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
  \to \Set/O$ that is nautrally isomorphic to the extension of a
  dependent polynomial from $I$ to $O$.
\end{definition}
With this interpretation of $F$ as a polynomial functor, $\el(F)$ and
$\map(F)$ are just interpreted as the action of the functor on objects
and morphisms, respectively. We interpret the constructors
$\mathsf{K},\Var$, etc. on functors in the obvious way that matches
the definitional behavior of $\el$ and $\map$. The non-trivial part of
the construction is verifying that such such constructions are closed
under being polynomial. The details are tedious but straightforward
extension of prior work on dependent polynomials and indexed
containers and we have verified the construction in Agda.

We use dependent polynomials functors on sets as these are guaranteed
to have initial algebras \cite{initial-algebras-of-dep-poly}.  Further
these initial algebras are readily constructed in our Agda
implementation as an inductive type of $IW$ trees which are already
available in the cubical library of Agda \cite{iw,cubical-library}.
Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
$\sem{X}\gamma$-indexed family of polynomial functors from
$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
of these constructs a polynomial functor from $\sem{X}\gamma\times
\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
initial algebra of this functor, and the initial algebra structure is
used to interpret $\roll, \fold$ and the corresponding axioms.

The remaining details of the interpretation of linear terms as parse
transformers and verification of the equational axioms is a relatively
straightforward extension of existing semantics of linear logic in
monoidal categories, and is included in the
appendix\cite{linearLogicMonoidalCategory}.

\subsection{Agda Implementation}

This denotational semantics in grammars and parse transformers serves
as the basis for our prototype implementation of \theoryabbv in
cubical Agda. The implementation is a shallow embedding, meaning that
rather than formalizing a syntax of \theoryabbv types and terms, we
work directly with Agda types and a definition of a formal grammar as
a function $\String \to \Set$. Then we implement each of the type and
term constructors of \theoryabbv as combinators on formal grammars or
parse transformers, and use cubical Agda's equality type to model the
term equalities. Cubical Agda is convenient for this purpose as it has
built-in support for function extensionality which we use
extensively. Axioms such as distributivity of $\oplus/\&$ and
disjointness of constructors are then provable directly in Agda, and
we are careful to only construct grammars, terms and proofs using
constructs that are possible in \theoryabbv. The main difference
between our shallow embedding and \theoryabbv is that our linear terms
are written in a combinator-style, without being able to use named
variables in linear terms. A benefit of this shallow embedding is that
the parsers are immediately available to a larger Agda development, as
they are just normal Agda code. In future work we will look to
implement a type checker for a syntax closer to the presentation in
this paper, but with the goal of still making it integrate with an
existing proof assistant.

@@

\section{Discussion and Future Work}
\label{sec:discussion}
\subsection{Typed Approaches to Grammars}

\steven{Talk about Conal Elliott's Automatic and Symbolic differentiation paper}
\steven{Talk briefly about this recent arxiv paper? https://arxiv.org/abs/2411.06358}

\subsubsection{Lambek Calculus and Categorial Grammar}
